Why is Stochastic Lambda > Deterministic?

### 2022-04-22 notes/understanding stochastic lambda.Rmd
tweak and render notes

---
title: "Why is Stochastic Lambda > Deterministic?"
author: "Eric R. Scott"
date: "2022-04-21"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    number_sections: yes
    highlight: kate
    theme:
      version: 4
      bootswatch: flatly
---

@@

conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
```

@@

```{r warning=FALSE}
lambdas <- data_full %>%
  filter(year>1998) %>% 
  group_by(year_fct) %>% 
  group_split() %>% 
  map_dbl(~{
    vit_list_det_ff = c(
      list(
        vit_surv = surv_det(.x, habitat = "1-ha"),
        vit_size = size_det(.x, habitat = "1-ha"),
        vit_flwr = flwr_det(.x, habitat = "1-ha"),
        vit_size_sdlg = size_sdlg_det(.x, habitat = "1-ha"),
        vit_surv_sdlg = surv_sdlg_det(.x, habitat = "1-ha")
      ),
      vit_other_ff)

    make_proto_ipm_det(vit_list_det_ff) %>% 
      make_ipm(iterations = 1000,  #only needs 100 to converge
               normalize_pop_size = FALSE, # to run as PVA
               usr_funs = list(get_scat_params = get_scat_params)) %>% lambda()
  })
worst_year <- c(1999:max(data_full$year))[which.min(lambdas)]; min(lambdas)
best_year <- c(1999:max(data_full$year))[which.max(lambdas)]; max(lambdas)

@@

```{r}
vit_list_det_ff = c(
  list(
    vit_surv = surv_det(two_yrs, habitat = "1-ha"),
    vit_size = size_det(two_yrs, habitat = "1-ha"),
    vit_flwr = flwr_det(two_yrs, habitat = "1-ha"),
    vit_size_sdlg = size_sdlg_det(two_yrs, habitat = "1-ha"),
    vit_surv_sdlg = surv_sdlg_det(two_yrs, habitat = "1-ha")
  ),
  vit_other_ff)

@@

The stochastic IPM requires re-writing the proto_ipm function because it's now using different parameter sets for year.

```{r}
## forets fragments
vit_list_stoch_ff = c(
  list(
    vit_surv = surv_raneff(two_yrs, habitat = "1-ha"),
    vit_size = size_raneff(two_yrs, habitat = "1-ha"),
    vit_flwr = flwr_raneff(two_yrs, habitat = "1-ha"),
    vit_size_sdlg = size_sdlg_raneff(two_yrs, habitat = "1-ha"),
    vit_surv_sdlg = surv_sdlg_raneff(two_yrs, habitat = "1-ha")
  ), 
  vit_other_ff)

@@

Iterate IPM

```{r}
ipm_stoch_ff <- 
  proto_ipm_stoch_ff %>%
  make_ipm(iterations = 1000,
           normalize_pop_size = FALSE, # to run as PVA
           usr_funs = list(get_scat_params = get_scat_params))
```

@@

```{r}
lambda(ipm_det_ff)
lambda(ipm_stoch_ff, log = FALSE)
```

Yes, stochastic lambda is still higher!

@@

```{r}
all_lams <- tibble(year = ipm_stoch_ff$env_seq,
       lambda = lambda(ipm_stoch_ff, type_lambda = "all")[,1])
all_lams %>% group_by(year) %>% summarize(mean(lambda)) %>% knitr::kable()
```

@@

@@

```{r}
newdata <- expand.grid(
  log_size_prev = seq(min(two_yrs$log_size_prev, na.rm = TRUE), max(two_yrs$log_size_prev, na.rm = TRUE), length.out = 50),
  year_fct = as.character(c(2001, 2003))
)

@@

det_fitted <- 
  vit_list_det_ff %>% 
  map_df(~safe_predict(.x, newdata = newdata, type = "response")) %>% 
  bind_cols(newdata) %>%
  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
  mutate(year_fct = "deterministic")

stoch_fitted <- 
  vit_list_stoch_ff %>%
  map_df(~safe_predict(.x, newdata= newdata, type = "response")) %>%
  bind_cols(newdata) %>%
  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_")

bind_rows(stoch_fitted, det_fitted) %>% 
  ggplot(aes(x = log_size_prev, y = .fitted, color = year_fct)) +
  geom_line() +
  facet_wrap(~vital_rate, scales = "free")

@@

Hold on.
Those numbers for size_sdlg don't make sense!
The mean size a seedling establishes at should be more like 2

```{r}
two_yrs %>% filter(sdlg_prev == TRUE, !is.na(log_size), habitat== "1-ha") %>% select(-L, -spei_history) %>% pull(log_size) %>% hist()
data_full %>% filter(ha_id_number == "2") %>% View()
data_full %>% filter(sdlg_prev == TRUE, !is.na(log_size), habitat == "1-ha") %>% View()
```

>>>>

---
title: "Why is Stochastic Lambda > Deterministic?"
author: "Eric R. Scott"
date: "2022-04-21"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    number_sections: yes
    highlight: kate
    code_folding: show
    theme:
      version: 4
      bootswatch: flatly
---

@@

conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
set.seed(123)
```

@@

```{r warning=FALSE, class.source = 'fold-hide'}
single_years <-   
  data_full %>%
  filter(year>1999) %>% 
  select(-L, -spei_history) %>% 
  group_by(year_fct) %>% 
  group_split()

vit_list_yr <- 
  single_years %>% 
  map(~{
    vit_list_det_ff = c(
      list(
        vit_surv = surv_det(.x, habitat = "1-ha"),
        vit_size = size_det(.x, habitat = "1-ha"),
        vit_flwr = flwr_det(.x, habitat = "1-ha"),
        vit_size_sdlg = size_sdlg_det(.x, habitat = "1-ha"),
        vit_surv_sdlg = surv_sdlg_det(.x, habitat = "1-ha")
      ),
      vit_other_ff)
    })

ipms_single_yr <-
  map(vit_list_yr, ~{
    make_proto_ipm_det(.x) %>% 
      make_ipm(iterations = 1000,  #only needs 100 to converge
               normalize_pop_size = FALSE, # to run as PVA
               usr_funs = list(get_scat_params = get_scat_params))
  })

lambdas <- ipms_single_yr %>% map_dbl(lambda)
```

```{r warning=FALSE}
#worst year
(worst_year <- c(2000:max(data_full$year))[which.min(lambdas)]); min(lambdas)

#best year
(best_year <- c(2000:max(data_full$year))[which.max(lambdas)]); max(lambdas)

@@

```{r warning=FALSE, class.source = 'fold-hide'}
vit_list_det_ff = c(
  list(
    vit_surv = surv_det(two_yrs, habitat = "1-ha"),
    vit_size = size_det(two_yrs, habitat = "1-ha"),
    vit_flwr = flwr_det(two_yrs, habitat = "1-ha"),
    vit_size_sdlg = size_sdlg_det(two_yrs, habitat = "1-ha"),
    vit_surv_sdlg = surv_sdlg_det(two_yrs, habitat = "1-ha")
  ),
  vit_other_ff)

@@

The stochastic IPM requires re-writing the `proto_ipm` because it's now using different parameter sets for year.
This now uses vital rates models where there is a random effect of year fit such that each year can have a different shape smoother (but not different wiggliness).

```{r warning=FALSE, class.source = 'fold-hide'}
## forets fragments
vit_list_stoch_ff = c(
  list(
    vit_surv = surv_raneff(two_yrs, habitat = "1-ha"),
    vit_size = size_raneff(two_yrs, habitat = "1-ha"),
    vit_flwr = flwr_raneff(two_yrs, habitat = "1-ha"),
    vit_size_sdlg = size_sdlg_raneff(two_yrs, habitat = "1-ha"),
    vit_surv_sdlg = surv_sdlg_raneff(two_yrs, habitat = "1-ha")
  ), 
  vit_other_ff)

@@

Iterate IPM with 1000 iterations

```{r warning=FALSE}
ipm_stoch_ff <- 
  proto_ipm_stoch_ff %>%
  make_ipm(iterations = 1000,
           normalize_pop_size = FALSE, # to run as PVA
           usr_funs = list(get_scat_params = get_scat_params))
```

@@

```{r}
lambda(ipm_det_ff) 
lambda(ipm_stoch_ff, log = FALSE, burn_in = 0.2)
```

Yes, stochastic lambda is still a bit higher!

@@

```{r}
all_lams <- 
  tibble(year = ipm_stoch_ff$env_seq,
         lambda = lambda(ipm_stoch_ff, type_lambda = "all")[,1])
all_lams %>%
  group_by(year) %>%
  summarize(mean(lambda))
```

@@

How do the mean lambdas from the good and bad year compare to the deterministic lambdas from individual years?

```{r}
max(lambdas)
min(lambdas)
```

The deterministic lambda for 2001 is a lot higher than the mean of the 2001 lambdas from the stochastic simulation and the deterministic lambda for 2003 is a lot lower than the mean of the 2003 lambdas from the stochastic simulation.
This makes sense to me, as fitting year as a random effect rather than a fixed effect will shrink estimates toward a global mean.

@@

```{r class.source = 'fold-hide'}
newdata <- expand.grid(
  log_size_prev = seq(min(two_yrs$log_size_prev, na.rm = TRUE), max(two_yrs$log_size_prev, na.rm = TRUE), length.out = 50),
  year_fct = as.character(c(2001, 2003))
)

@@

#deterministic, just 2001
det_fitted_2001 <- 
  vit_list_yr[[which.max(lambdas)]] %>% 
  map_df(~safe_predict(.x, newdata = newdata, type = "response")) %>% 
  bind_cols(newdata) %>% 
  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
  mutate(year_fct = "det. 2001")

#deterministic, just 2003
det_fitted_2003 <- 
  vit_list_yr[[which.min(lambdas)]] %>% 
  map_df(~safe_predict(.x, newdata = newdata, type = "response")) %>% 
  bind_cols(newdata) %>% 
  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
  mutate(year_fct = "det. 2003")

#deterministic with both years combined
det_fitted <- 
  vit_list_det_ff %>% 
  map_df(~safe_predict(.x, newdata = newdata, type = "response")) %>% 
  bind_cols(newdata) %>%
  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
  mutate(year_fct = "det. both years")

#stochastic
stoch_fitted <- 
  vit_list_stoch_ff %>%
  map_df(~safe_predict(.x, newdata= newdata, type = "response")) %>%
  bind_cols(newdata) %>%
  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
  mutate(year_fct = paste("stoch.", year_fct))

bind_rows(stoch_fitted, det_fitted, det_fitted_2001, det_fitted_2003) %>% 
  ggplot(aes(x = log_size_prev, y = .fitted, color = year_fct, linetype = year_fct)) +
  geom_line(alpha = 0.9) +
  facet_wrap(~vital_rate, scales = "free")

@@

For the stochastic models, the two years are nearly indistinguishable for seedlings.
I believe this is because of shrinkage---there is little information for either year so means are shrunk toward global average.
You can see this more clearly for flowering probability where the order of lines at the larger end is 2001 deterministic (essentially a fixed effect), 2001 stochastic (random effect), both years combined, 2003 stochastic (random effect), and 2003 deterministic (fixed effect).
